Challenges in Developing Great Quasi-Monte Carlo Software

### 2023-07-12 MCQMC2022ChallengesQMCSoftwareArticle_revision_1.tex
SC: fix a bib item
Fred 2023 Mac 14 Updates

\abstract{
Quasi-Monte Carlo (QMC) methods have developed over several decades. With the explosion in computational science, there is a need for great software that implements QMC algorithms. We summarize the  QMC software that has been developed to date, propose some criteria for developing great QMC software, and suggest some steps toward achieving great software.  We illustrate these criteria and steps in practice with the Quasi-Monte Carlo Python library (QMCPy), an open-source community software framework, extensible by design with common programming interfaces to an increasing number of existing or emerging QMC libraries developed by the greater community of QMC researchers. 
}

@@

We draw on our collective experiences as
members of the academic QMC community, developers of open-source and commercial scientific software, and users of a wide range of scientific software as research scientists or data scientists. Many of the insights that we have gained have come through our development of the Guaranteed Automatic Integration Library (GAIL)~\SCComment{\cite{TonEtAl22a}} in MATLAB and the Quasi-Monte Carlo Python library (ChoEtal22a)~\SCComment{\cite{QMCPy2020a}}.

@@

Great QMC software opens up new application areas for QMC methods by allowing practitioners to compare new methods to existing ones.  QMC software has been successful in quantitative finance, uncertainty quantification, and image rendering. Unexpectedly good or bad computational results lead to open theoretical questions.  For example, the early application of QMC to high dimensional integrals arising in financial risk \cite{PasTra95} led to a wave of theoretical results on the tractability of integration in weighted spaces \cite{Woz99a,DicEtal14a,NovWoz10a}.

@@

We illustrate in Figure~\ref{CDHJSfigbeam} how UM-Bridge connects QMCPy with a differential equation boundary value problem solver for a cantilevered beam \SCComment{of length~$L$}, with a random Young's modulus \SCComment{$m(x)$, assumed to be piecewise constant over three regions, each of which with a uniform distribution of uncertainty~\cite{ParSee22a}, i.e., $m(x) = \left[\begin{matrix} m_1(x), m_2(x), m_3(x) \end{matrix} \right]$ with $m_i(x) \sim \calU[1,1.2]^3$.  The beam position $x$ is discretized at $N=31$ equidistant points, and 
the vertical deflection $u(x)$ when the beam is subject to a given external load  $g(x)$ satisfies the Euler-Bernoulli partial differential equation (PDE):
$$ \frac{\partial^2}{\partial x^2}\left[ \exp(m(x)) \frac{\partial^2  u(x)}{\partial x^2}\right] = g(x),$$
%where $m(x) = \log E(x)$ is the log of an effective stiffness $E(x)$.  The  
with the following boundary conditions:
$$ \left.u(x)\right|_{x=0} = 0, \qquad 
\left.\frac{\partial u(x)}{\partial x}\right|_{x=0} = 0, \qquad
\left.\frac{\partial^2 u(x)}{\partial x^2}\right|_{x=L} = 0, \qquad  \left.\frac{\partial^3 u(x)}{\partial x^3}\right|_{x=L} = 0. $$
This PDE can be discretized with finite differences yielding a linear system $K(m) \hat{u} = \hat{g}$, where $\hat{u}$ and $\hat{g} \in \mathbb{R}^N$. }
%approximates $u(x)$ and $m(x)$ at the nodes, respectively. 
%The forward model $F:\mathbb{R}^3 \rightarrow \mathbb{R}^N $ maps $m = \left[\begin{matrix}m_1, m_2, m_3 \end{matrix}\right]^T$ to $\hat{u}$, while $Q : \mathbb{R}^N \rightarrow \mathbb{R}^2 $ selects the estimated solution at the 10th and 25th nodes, i.e., $ Q(\hat{u}) := \left[\begin{matrix} \hat{u}_{10}, \hat{u}_{25} \end{matrix} \right]^T$. The objective is to determine the distributions of $Q(F(m))$. 
\LSComment{The beam model was written in C++ using the MUQ's model framework~\cite{MUQ}. It is one of the UM-Bridge's benchmark problem library, available as a container.}
QMCPy simulates the random Young's modulus and computes the expected deflection of the beam at the node points for a sequence of error tolerances using lattice rules.  By taking advantage of \LSComment{clusters and containers}, UM-Bridge provides automatic load balancing to facilitate seamless parallel model evaluation. \LSComment{As a result, thread-parallel QMCPy can offload costly model evaluations to remote clusters~\cite{seelinger2023lowering}.}

>>>>

\abstract{
Quasi-Monte Carlo (QMC) methods have developed over several decades. With the explosion in computational science, there is a need for great software that implements QMC algorithms. We summarize the  QMC software that has been developed to date, propose some criteria for developing great QMC software, and suggest some steps toward achieving great software.  We illustrate these criteria and steps with the Quasi-Monte Carlo Python library (QMCPy), an open-source community software framework, extensible by design with common programming interfaces to an increasing number of existing or emerging QMC libraries developed by the greater community of QMC researchers.
}

@@

We draw on our collective experiences as
members of the academic QMC community, developers of open-source and commercial scientific software, and users of a wide range of scientific software as research scientists or data scientists. Many of the insights that we have gained have come through our development of the Guaranteed Automatic Integration Library (GAIL)~\cite{TonEtAl22a} in MATLAB and the Quasi-Monte Carlo Python library (QMCPy)~\cite{QMCPy2020a}.

@@

Great QMC software opens up new application areas for QMC methods by allowing practitioners to compare new methods to existing ones.  QMC software has been successful in quantitative finance, uncertainty quantification, and image rendering. Unexpectedly good or bad computational results prompt theoretical questions.  For example, the early application of QMC to high dimensional integrals arising in financial risk \cite{PasTra95} led to a wave of theoretical results on the tractability of integration in weighted spaces \cite{Woz99a,DicEtal14a,NovWoz10a}.

@@

We illustrate in Figure~\ref{CDHJSfigbeam} how UM-Bridge connects QMCPy with a differential equation boundary value problem solver for a cantilevered beam \SCComment{of length~$L$}, with a random Young's modulus \SCComment{$m(x)$, assumed to be piecewise constant over three regions, each of which with a uniform distribution of uncertainty~\cite{ParSee22a}, i.e., $m(x) = \left[\begin{matrix} m_1(x), m_2(x), m_3(x) \end{matrix} \right]$ with $m_i(x) \sim \calU[1,1.2]^3$.  The beam position $x$ is discretized at $N=31$ equidistant points, and
the vertical deflection $u(x)$ when the beam is subject to a given external load  $g(x)$ satisfies the Euler-Bernoulli partial differential equation (PDE):
$$ \frac{\partial^2}{\partial x^2}\left[ \exp(m(x)) \frac{\partial^2  u(x)}{\partial x^2}\right] = g(x),$$
%where $m(x) = \log E(x)$ is the log of an effective stiffness $E(x)$.  The
with the following boundary conditions:
$$ \left.u(x)\right|_{x=0} = 0, \qquad
\left.\frac{\partial u(x)}{\partial x}\right|_{x=0} = 0, \qquad
\left.\frac{\partial^2 u(x)}{\partial x^2}\right|_{x=L} = 0, \qquad  \left.\frac{\partial^3 u(x)}{\partial x^3}\right|_{x=L} = 0. $$
This PDE can be discretized with finite differences yielding a linear system $K(m) \hat{u} = \hat{g}$, where $\hat{u}$ and $\hat{g} \in \mathbb{R}^N$. }
%approximates $u(x)$ and $m(x)$ at the nodes, respectively.
%The forward model $F:\mathbb{R}^3 \rightarrow \mathbb{R}^N $ maps $m = \left[\begin{matrix}m_1, m_2, m_3 \end{matrix}\right]^T$ to $\hat{u}$, while $Q : \mathbb{R}^N \rightarrow \mathbb{R}^2 $ selects the estimated solution at the 10th and 25th nodes, i.e., $ Q(\hat{u}) := \left[\begin{matrix} \hat{u}_{10}, \hat{u}_{25} \end{matrix} \right]^T$. The objective is to determine the distributions of $Q(F(m))$.
\LSComment{The beam model was written in C++ using the MUQ's model framework~\cite{MUQ2}. It is one of the UM-Bridge's benchmark problem library, available as a container.}
QMCPy simulates the random Young's modulus and computes the expected deflection of the beam at the node points for a sequence of error tolerances using lattice rules.  By taking advantage of \LSComment{clusters and containers}, UM-Bridge provides automatic load balancing to facilitate seamless parallel model evaluation. \LSComment{As a result, thread-parallel QMCPy can offload costly model evaluations to remote clusters~\cite{seelinger2023lowering}.}
